Predicting Ensemble Location and Width in Binaural Recordings of Music with Convolutional Neural Networks

### 2024-06-19 paper/main.tex
Add experiments with noise prediction

\begin{abstract}
  Binaural audio technology has been in existence for many years, but its popularity has significantly increased over the past decade as a consequence of advancements in virtual reality and streaming technologies. Along with its growing popularity, the quantity of publicly accessible binaural audio materials has also expanded. Consequently, there is now a need for automated and objective measurements of spatial content information, with ensemble location and width being the most important. This study presents a novel method for predicting ensemble location and width in binaural recordings. To this end, 30 head-related transfer functions and 192 binaural music recordings from publicly accessible multi-track recording repositories were used to synthesize 23,040 binaural recordings. The synthesized recordings were then used to train a multi-task convolutional neural network model, the aim of which was to predict the location and width of the ensemble for unseen recordings. The results indicate that models of ensemble location and width can be successfully constructed with low prediction errors --- $4.76\degree$ ($\pm0.10\degree$) for ensemble location and $8.57\degree$ ($\pm0.19\degree$) for ensemble width. This approach is a novel method for predicting both ensemble location and width, offering a more accurate representation of spatial properties. This suggests significant potential for advancing spatial audio applications in virtual reality and streaming technologies, by providing audio engineers with tools that can leverage this method to enhance spatial audio experiences.
\end{abstract}

@@

The advancement of sophisticated machine learning techniques, particularly deep learning networks, has prompted an intriguing exploration of the extent to which these tools can emulate the human auditory system without relying on advanced spatial audio feature engineering, traditionally employed in audio source localization tasks \cite{yang_deepear_2022, vera-diaz_towards_2018, pang_multitask_2019}. To investigate this, the study developed an audio localization method based on a multi-task convolutional neural network (CNN) model.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{../figures/samples.pdf}
  \caption{\label{fig:samples}Example magnitude spectrograms from randomly selected synthesized binaural music recordings, used as input data for training and evaluating the convolutional neural network (CNN) in this study. }
\end{figure}

Prior to input into the model, the binaural recordings of music were transformed into magnitude spectrograms. Figure \ref{fig:samples} displays example spectrograms from four distinct binaural recordings, exhibiting a broad bandwidth and the diversity of the signals used in this study. It is important to note that the original spectrograms were preserved in a floating-point precision matrix format, which prevented any loss of information due to precision conversion. The spectrograms were calculated using the Fast Fourier Transform (FFT) algorithm, with a limitation of 150 bands, spaced linearly from 100 Hz to 16 kHz. Additionally, a Hamming window of 40 ms was applied to each frame of the signal, resulting in a total of 349 time frames. This procedure was conducted for both the left and right channels, yielding two spectrograms for each binaural sample. Each sample was represented by a matrix of dimensions $2 \times 349 \times 150$. This method parallels the procedure presented in \cite{zielinski_automatic_2022}.

@@

@@

The mean absolute prediction error for ensemble width was $8.57^\circ$ ($\pm0.19^\circ$), while the error for ensemble location prediction was $4.76^\circ$ ($\pm0.10^\circ$). Although both ensemble parameters were constrained within the same angular span of $90^\circ$, the ensemble location was predicted with significantly better precision --- approximately $80\%$ better. This shows that the model prediction quality is much better for ensemble location prediction than width prediction. This difference can be attributed to the fact that the model's performance on ensemble width deteriorates significantly for wider ensembles, whereas location prediction remains largely unaffected. For a visual comparison, please refer to Figures \ref{fig:mae_location} and \ref{fig:mae_width}.

@@

\clearpage

\section*{Appendix A}
\label{appendix:a_hrtf}

>>>>

\begin{abstract}
  Binaural audio technology has been in existence for many years, but its popularity has significantly increased over the past decade as a consequence of advancements in virtual reality and streaming technologies. Along with its growing popularity, the quantity of publicly accessible binaural audio materials has also expanded. Consequently, there is now a need for automated and objective measurements of spatial content information, with ensemble location and width being the most prominent. This study presents a novel method for predicting ensemble location and width in binaural recordings. To this end, 30 head-related transfer functions and 192 binaural music recordings from publicly accessible multi-track recording repositories were used to synthesize 23,040 binaural recordings. The synthesized recordings were then used to train a multi-task convolutional neural network model, the aim of which was to predict the location and width of the ensemble for unseen recordings. The results indicate that models of ensemble location and width can be successfully constructed with low prediction errors --- $4.76\degree$ ($\pm0.10\degree$) for ensemble location and $8.57\degree$ ($\pm0.19\degree$) for ensemble width. The method developed in this study outperforms previous spatiogram-based methods recently published in the literature and holds promise for future development as a part of a novel tool for audio engineers to assess binaural audio.
\end{abstract}

@@

The advancement of sophisticated machine learning techniques, particularly deep learning networks, has prompted an intriguing exploration of the extent to which these tools can emulate the human auditory system without relying on advanced spatial audio feature engineering, traditionally employed in audio source localization tasks \cite{yang_deepear_2022, vera-diaz_towards_2018, pang_multitask_2019}. To investigate this, the study developed an audio localization method based on a multi-task convolutional neural network (CNN) model \cite{lecun_handwritten_1989, lecun_deep_2015}.

@@

% Zamiast tego dac obrazek z glowami i spektrogramami, lewy i prawi i roznica, sprawdzic czy to ma sens, potencjalnie usunac
% \begin{figure}[ht]
%   \centering
%   \includegraphics[width=\linewidth]{../figures/samples.pdf}
%   \caption{\label{fig:samples}Example magnitude spectrograms from randomly selected synthesized binaural music recordings, used as input data for training and evaluating the convolutional neural network (CNN) in this study. }
% \end{figure}

% Figure \ref{fig:samples} displays example spectrograms from four distinct binaural recordings, exhibiting a broad bandwidth and the diversity of the signals used in this study

Prior to input into the model, the binaural recordings of music were transformed into magnitude spectrograms. It is important to note that the original spectrograms were preserved in a floating-point precision matrix format, which prevented any loss of information due to precision conversion. The spectrograms were calculated using the Fast Fourier Transform (FFT) algorithm, with a limitation of 150 bands, spaced linearly from 100 Hz to 16 kHz. Additionally, a Hamming window of 40 ms was applied to each frame of the signal, resulting in a total of 349 time frames. This procedure was conducted for both the left and right channels, yielding two spectrograms for each binaural sample. Each sample was represented by a matrix of dimensions $2 \times 349 \times 150$. This method parallels the procedure presented in \cite{zielinski_automatic_2022}.

@@

The network topology used in this study was highly inspired by a AlexNet convolutional neural network \cite{krizhevsky_imagenet_2012}. Although AlexNet network was designed for image classification, the binaural recordings in this study were converted into magnitude spectrograms, as described in Section \ref{subsec:feature_extraction}. The network is consisting of a series of convolutional units followed by a series of classification units.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{../pictures/architecture.pdf}
  \caption{\label{fig:architecture}Topology of the Convolutional Neural Network (CNN) used for identifying ensemble location and width}
\end{figure}

@@

% jak uzyskane wyniki koresponduja z literatura: w jjednym przypadku ejst lepsza niz metoda oparta na spatiogramach, a w drugim przypadku nie ma 
The mean absolute prediction error for ensemble width was $8.57^\circ$ ($\pm0.19^\circ$), while the error for ensemble location prediction was $4.76^\circ$ ($\pm0.10^\circ$). Although both ensemble parameters were constrained within the same angular span of $90^\circ$, the ensemble location was predicted with significantly better precision --- approximately $80\%$ better. This shows that the model prediction quality is much better for ensemble location prediction than width prediction. This difference can be attributed to the fact that the model's performance on ensemble width deteriorates significantly for wider ensembles, whereas location prediction remains largely unaffected. For a visual comparison, please refer to Figures \ref{fig:mae_location} and \ref{fig:mae_width}.

@@

% 1. napisac wnioski, potem occulusion. we wnioskach odniesc sie do spatiogramach

\clearpage
\section*{Appendix A}
\label{appendix:a_hrtf}
